Factors affecting reliablity of state-space age-structured assessment models

### 2025-01-16 Project_0/manuscript/manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1; Andrew Applegate; Greg Britten; Elizabeth N. Brooks; Gavin Fay; Alex Hansell; Christopher M. Legault; Brandon Muffley; Brian C. Stock; John Wiedenmann},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
%\doublespacing
\usepackage{lineno}
%\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
\newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
\newcommand{\ajms}{African Journal of Marine Science\xspace}
\newcommand{\amb}{Advances in Marine Biology\xspace}
\newcommand{\bms}{Bulletin of Marine Science\xspace}
\newcommand{\bjssf}{Bulletin of the Japanese Society of Scientific Fisheries\xspace}
\newcommand{\cb}{Conservation Biology\xspace}
\newcommand{\cjfas}{Canadian Journal of Fisheries and Aquatic Sciences\xspace}
\newcommand{\ea}{Ecological Applications\xspace}
\newcommand{\eer}{Evolutionary Ecology Research\xspace}
\newcommand{\elet}{Ecology Letters\xspace}
\newcommand{\emod}{Ecological Modelling\xspace}
\newcommand{\ebf}{Environmental Biology of Fishes\xspace}
\newcommand{\ff}{Fish and Fisheries\xspace}
\newcommand{\fo}{Fisheries Oceanography\xspace}
\newcommand{\fr}{Fisheries Research\xspace}
\newcommand{\fb}{Fishery Bulletin\xspace}
\newcommand{\ijms}{ICES Journal of Marine Science\xspace}
\newcommand{\iccat}{Collective Volume of Scientific Papers ICCAT\xspace}
\newcommand{\jae}{Journal of Animal Ecology\xspace}
\newcommand{\jai}{Journal of Applied Ichthyology\xspace}
\newcommand{\jdc}{Journal Du Conseil International Pour L'exploration De La Mer\xspace}
\newcommand{\jdcp}{Journal Du Conseil Permanent International Pour L'exploration De La Mer\xspace}
\newcommand{\jembe}{Journal of Experimental Marine Biology and Ecology\xspace}
\newcommand{\jfb}{Journal of Fish Biology\xspace}
\newcommand{\jsr}{Journal of Sea Research\xspace}
\newcommand{\jtb}{Journal of Theoretical Biology\xspace}
\newcommand{\jfrbc}{Journal of the Fisheries Research Board of Canada\xspace}
\newcommand{\jnwafs}{Journal of Northwest Atlantic Fisheries Science\xspace}
\newcommand{\mcf}{Marine and Coastal Fisheries: Dynamics, Management, and Ecosystem Science\xspace}
\newcommand{\mb}{Marine Biology\xspace}
\newcommand{\meps}{Marine Ecology Progress Series\xspace}
\newcommand{\mfr}{Marine Fisheries Review\xspace}
\newcommand{\mpb}{Marine Pollution Bulletin\xspace}
\newcommand{\najfm}{North American Journal of Fisheries Management\xspace}
\newcommand{\nzjmfr}{New Zealand Journal of Marine and Freshwater Research\xspace}
\newcommand{\pnas}{Proceedings of the National Academy of Sciences USA\xspace}
\newcommand{\rpvrciemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Internationale pour l'Exploration de la Mer\xspace}
\newcommand{\rpvrcpiemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Permanent Internationale pour l'Exploration de la Mer\xspace}
\newcommand{\rfbf}{Reviews in Fish Biology and Fisheries\xspace}
\newcommand{\sajms}{South African Journal of Marine Science\xspace}
\newcommand{\tafs}{Transactions of the American Fisheries Society\xspace}

@@

\usepackage{xspace}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{makecell}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
\renewcommand\theadfont{}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{cjfas2.bst}

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1} \and Andrew
Applegate \and Greg Britten \and Elizabeth N. Brooks \and Gavin
Fay \and Alex Hansell \and Christopher M. Legault \and Brandon
Muffley \and Brian C. Stock \and John Wiedenmann}
\date{}

@@

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water
Street, Woods Hole, MA 02543 USA\\

@@

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within ICES, Canada, and the
Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}. State-space
approaches that use random effects to parameterize process errors is
considered best practice and a requirement for the next generation of
stock assessment models \citep{hoyleetal22, punt23}.

Much is known about the reliability of state-space models that are
linear or Gaussian \citep{aeberhardetal18}, but applications in
fisheries management are nonlinear and typically include multiple types
of observations with varying distributional assumptions. Furthermore,
there is a wide range of potential random effects structures and model
parameters that can be treated as random effects in assessment models.
We know relatively little about the factors affecting statistical
reliability of such models or the ability of information criteria to
distinguish among such alternative structures.

@@

We used the Woods Hole Assessment Model (WHAM) to configure operating
and EMs in our simulation study \citep[]{millerstock20,stockmiller21}.
WHAM is an R package freely available as a github repository. For this
study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This packages has also been used to
configure operating and EMs for closed loop simulations evaluating
index-based assessment methods \citep{legaultetal23} and is used for
management of haddock, butterfish, American plaice, bluefish, Atlantic
cod, and black sea bass in the Northeast US.

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects are assumed:
abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
(R+Sel), or index catchability (R+q). For each OM assumptions about
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from recent
applications of WHAM in management of stocks of haddock, butterfish, and
American plaice in the NE US.

@@

We did not use the bias-correction feature for process errors or
observations described by \citet{stockmiller21} for operating and EMs.
Simulations and model fitting were all carried out on the University of
Massachusetts Green High-Performance Computing Cluster. All code we used
to perform the simulation study and summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population model tracks 10 age classes: ages 1 to 10+ and we assume
spawning occurs 1/4 of the way through the year. The maturity at age was
a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88 and assumed
known in all EMs (Figure \ref{om_maturity}).

Weight at age was generated with a LVB growth function \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
relationship such that \[
W_a = \theta_1 L_a^{\theta_2}
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
\ref{om_waa}).

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all OMs. All post-recruit
productivity components are constant in the NAA and survey catchability
process error OMs. Therefore steepness and unfished recruitment are also
constant over the time period for those OMs \citep{millerbrooks21}. We
specified unfished recruitment = \(R_0 = e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean rate was 0.2.

Two alternative fishing histories were used for OMs. In the first
scenario, the stock experiences overfishing for the first 20 years and
fishing at \Fmsy for the last 20 years. In the second scenario, the
stock is fished at \Fmsy for the entire time period. The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NE groundfish stocks from \citet{wiedenmannetal19}.
\Citet{legaultetal23} also used similar approaches to defining fishing
mortality histories for OMs.

Initial population was configured at the equilibrium distribution
fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\) for the two
alternative fishing histories. That is for a deterministic model, the
age composition would not change over time when the fishing mortality
was constant at the respective level.

For OMs with time-varying random effects for M, steepness is not
constant, but we used the same alpha and beta parameters as other OMs
this equates to a steepness and R0 at the mean of the time series
process for M. For OMs with time-varying random effects for fishery
selectivity, \Fmsy is also not constant however we use the same F
history as other OMs which corresponds to Fmsy at the mean selectivity
parameters.

@@

Two time series of surveys are assumed and observed in numbers rather
than biomass for the entire 40 year period with one occurring in the
spring (0.25 way through the year) and one in the fall (0.75 way through
the year). Actually we have it currently configured that both occur 0.5
way through the year. Catchability of both surveys are assumed to be
0.1. Like the fishing fleet, we assumed logistic selectivity for both
indices with \(a_{50} = 5\) and slope = 1 and a logistic-normal
distribution for the age-composition observations.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance assumptions for recruitment and
survival results in 24 different OMs with random effects on recruitment
and survival (Table \ref{naa_om_table}).

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
natural mortality results in 16 different OMs with random effects on
natural mortality (Table \ref{M_om_table}). Internally, WHAM treats
natural mortality as a log-transformed parameter \[
\log M_y = \mu_M + \epsilon_{M,y}
\] that is a linear combination of a mean \(\mu_M\) and any annual
random effects marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The values
for the marginal variance and correlation parameters assumed in the OMs
are given in Table \ref{M_om_table}.

@@

The factorial combinations of fishing histories, observation error
assumptions, and marginal variance and autocorrelation assumptions for
selectivity parameters results in 16 different OMs with random effects
on selectivity (Table \ref{sel_om_table}). Internally, WHAM treats
selectivity parameter \(s\) as a logit-transformed parameter \[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters are either \(a_50\) or
slope parameters and we assume bounds of 0 and 10 for all selectivity
parameters for all operating and EMs. The values for the marginal
variance and correlation parameters assumed in the OMs are given in
Table \ref{sel_om_table}.

@@

\bibliography{paper}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

\newcommand{\Fmsy}{\ensuremath{F_{\text{MSY}}}\xspace}
\newcommand{\Fspr}[1]{\ensuremath{F_{\text{{#1}\%}}}\xspace}
\newcommand{\afrb}{Alaska Fishery Research Bulletin\xspace}
\newcommand{\ajms}{African Journal of Marine Science\xspace}
\newcommand{\amb}{Advances in Marine Biology\xspace}
\newcommand{\bms}{Bulletin of Marine Science\xspace}
\newcommand{\bjssf}{Bulletin of the Japanese Society of Scientific Fisheries\xspace}
\newcommand{\cb}{Conservation Biology\xspace}
\newcommand{\cjfas}{Canadian Journal of Fisheries and Aquatic Sciences\xspace}
\newcommand{\ea}{Ecological Applications\xspace}
\newcommand{\eer}{Evolutionary Ecology Research\xspace}
\newcommand{\elet}{Ecology Letters\xspace}
\newcommand{\emod}{Ecological Modelling\xspace}
\newcommand{\ebf}{Environmental Biology of Fishes\xspace}
\newcommand{\ff}{Fish and Fisheries\xspace}
\newcommand{\fo}{Fisheries Oceanography\xspace}
\newcommand{\fr}{Fisheries Research\xspace}
\newcommand{\fb}{Fishery Bulletin\xspace}
\newcommand{\ijms}{ICES Journal of Marine Science\xspace}
\newcommand{\iccat}{Collective Volume of Scientific Papers ICCAT\xspace}
\newcommand{\jae}{Journal of Animal Ecology\xspace}
\newcommand{\jai}{Journal of Applied Ichthyology\xspace}
\newcommand{\jdc}{Journal Du Conseil International Pour L'exploration De La Mer\xspace}
\newcommand{\jdcp}{Journal Du Conseil Permanent International Pour L'exploration De La Mer\xspace}
\newcommand{\jembe}{Journal of Experimental Marine Biology and Ecology\xspace}
\newcommand{\jfb}{Journal of Fish Biology\xspace}
\newcommand{\jsr}{Journal of Sea Research\xspace}
\newcommand{\jtb}{Journal of Theoretical Biology\xspace}
\newcommand{\jfrbc}{Journal of the Fisheries Research Board of Canada\xspace}
\newcommand{\jnwafs}{Journal of Northwest Atlantic Fisheries Science\xspace}
\newcommand{\mcf}{Marine and Coastal Fisheries: Dynamics, Management, and Ecosystem Science\xspace}
\newcommand{\mb}{Marine Biology\xspace}
\newcommand{\meps}{Marine Ecology Progress Series\xspace}
\newcommand{\mfr}{Marine Fisheries Review\xspace}
\newcommand{\mpb}{Marine Pollution Bulletin\xspace}
\newcommand{\najfm}{North American Journal of Fisheries Management\xspace}
\newcommand{\nzjmfr}{New Zealand Journal of Marine and Freshwater Research\xspace}
\newcommand{\pnas}{Proceedings of the National Academy of Sciences USA\xspace}
\newcommand{\rpvrciemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Internationale pour l'Exploration de la Mer\xspace}
\newcommand{\rpvrcpiemm}{Rapports et Proc\`es-Verbaux des R\'eunions. Conseil Permanent Internationale pour l'Exploration de la Mer\xspace}
\newcommand{\rfbf}{Reviews in Fish Biology and Fisheries\xspace}
\newcommand{\sajms}{South African Journal of Marine Science\xspace}
\newcommand{\tafs}{Transactions of the American Fisheries Society\xspace}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Brandon
Muffley\textsuperscript{5} \and Brian C.
Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{15 January, 2025}

@@

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Woods Hole Oceanographic Institution\\
\(^4\)SMAST\\
\(^5\)Institute of Marine Research\\
\(^6\)Mid-Atlantic Fisheries Management Council\\
\(^7\)Rutgers University\\

@@

\hypertarget{abstract}{%
\subsection*{Abstract}\label{abstract}}
\addcontentsline{toc}{subsection}{Abstract}

Random effects can be included in state-space assessment models on many
processes and in many ways, and guidance is needed on statistical
reliability and model selection criteria. We simulated 72 operating
models with varying fishing history, observation error uncertainty, and
process error magnitude, correlation, and source (recruitment, survival,
fishery selectivity, catchability, and natural mortality). We fit
estimating models with different assumptions on the process error
source, whether (mean) natural mortality was estimated, and whether a
stock-recruit relationship was estimated.

Models that assumed the correct process error source had high
convergence and low bias. Bias was also low under most process error
assumptions when there was contrast in fishing pressure.
Stock-recruitment parameters were only reliably estimated in ideal
situations. Marginal AIC most accurately distinguished process errors on
recruitment, survival, and selectivity, as well as larger magnitude
process errors of other types. Retrospective patterns were generally
weak except for recruitment when observation error was high, even with
the correct process error assumptions. When models did exhibit some
retrospective pattern, estimating natural mortality often removed it.

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within ICES, Canada, and the
Northeast US (Nielsen and Berg 2014; Cadigan 2016; Pedersen and Berg
2017; Stock and Miller 2021). State-space approaches that use random
effects to parameterize process errors is considered best practice and a
requirement for the next generation of stock assessment models (Hoyle et
al. 2022; Punt 2023).

Much is known about the reliability of state-space models that are
linear or Gaussian (Aeberhard et al. 2018), but applications in
fisheries management are nonlinear and typically include multiple types
of observations with varying distributional assumptions. Furthermore,
there is a wide range of potential random effects structures and model
parameters that can be treated as random effects in assessment models.
We know relatively little about the factors affecting statistical
reliability of such models or the ability of information criteria to
distinguish among such alternative structures.

Li et al. (2024) investigated some aspects of inferences for operating
models with multiple sources of process error, but there are differences
for this paper.

@@

We used the Woods Hole Assessment Model (WHAM) to configure operating
and EMs in our simulation study (Miller and Stock 2020; Stock and Miller
2021). WHAM is an R package freely available as a github repository. For
this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This package has also been used to
configure operating and EMs for closed loop simulations evaluating
index-based assessment methods (Legault et al. 2023) and is used for
management of haddock, butterfish, American plaice, bluefish, Atlantic
cod, black sea bass, and yellowtail flounder in the Northeast US.

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects are assumed:
abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
(R+Sel), or index catchability (R+q). For each OM assumption about
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from
applications of WHAM in management of stocks of haddock, butterfish, and
American plaice in the NE US.

@@

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for operating and
EMs (Li et al. In review). Simulations and model fitting were all
carried out on the University of Massachusetts Green High-Performance
Computing Cluster. All code we used to perform the simulation study and
summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population consists of 10 age classes: ages 1 to 10+ and we assume
spawning occurs each year 1/4 of the way through the year. The maturity
at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
(Figure \ref{om_maturity}).

Weight at age was generated with a von Bertalanffy growth function \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
relationship such that \[
W_a = \theta_1 L_a^{\theta_2}
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
\ref{om_waa}).

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all OMs. All post-recruit
productivity components are constant in the NAA and survey catchability
process error OMs. Therefore steepness and unfished recruitment are also
constant over the time period for those OMs (Miller and Brooks 2021). We
specified unfished recruitment = \(R_0 = e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
\(\alpha=0.60\) and \(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without
process errors on natural mortality we assumed the rate was assumed 0.2.
For OMs with process errors on natural mortality the mean log natural
mortality rate was \(\log(0.2)\).

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years and fishing
at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)).
In the second scenario, the stock is fished at \Fmsy for the entire time
period. The magnitude of the overfishing assumptions is based on average
estimates of overfishing for NE groundfish stocks from (Wiedenmann et
al. 2019). Legault et al. (2023) also used similar approaches to
defining fishing mortality histories for OMs.

We specified initial population abundance at age at the equilibrium
distribution fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\)
for the two alternative fishing histories. This implies that, for a
deterministic model, the abundance at age would not change from the
first year to the next.

For OMs with time-varying random effects for M, steepness is not
constant, but we used the same alpha and beta parameters as other OMs
this equates to a steepness and R0 at the median of the time series
process for M. For OMs with time-varying random effects for fishery
selectivity, \Fmsy is also not constant however we use the same F
history as other OMs which corresponds to Fmsy at the mean selectivity
parameters.

@@

Two time series of surveys are assumed and observed in numbers rather
than biomass for the entire 40 year period with one occurring in the
spring (0.25 of each year) and one in the fall (0.75 of each year).
Catchability of both surveys are assumed to be 0.1. Like the fishing
fleet, we assumed logistic selectivity for both indices with
\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution for the
age-composition observations.

@@

For operating models with random effects on recruitment and(or) survival
(R, R+S) we assumed marginal standard deviations for recruitment of
\(\sigma_R \in \{0.5,1.5\}\) and marginal standard deviations for older
age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial
combination of these process error assumptions and fishing history (2
levels) and observation error (2 levels) scenarios described above
results in 24 different R (\(\sigma_{2+} = 0\)) and R+S operating models
(Table \ref{naa_om_table}).

@@

All R+M OMs treat natural mortality constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean that is constant across ages
\(\mu_{M} = \log(0.2)\) and any annual random effects marginally
distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
random effects were also included on recruitment with \(\sigma_R = 0.5\)
(hence, R+M). The marginal standard deviations we assumed for log
natural mortality random effects were \(\sigma_M \in \{0.1, 0.5\}\) and
AR1 autocorrelation parameters of \(\rho_M \in \{0,0.9\}\). The full
factorial combination of these process error assumptions and fishing
history (2 levels) and observation error (2 levels) scenarios described
above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

MORE SPECIFICS about correlation of random effects? Both selectivity
pars? just correlated by year? WHAM treats selectivity parameter \(s\)
as a logit-transformed parameter \[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters are either \(a_50\) or
slope parameters and we assume bounds of 0 and 10 for all selectivity
parameters for all operating and EMs. The marginal standard deviations
we assumed for logit scale random effects were
\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
of these process error assumptions and fishing history (2 levels) and
observation error (2 levels) scenarios described above results in 16
different R+Sel OMs (Table \ref{sel_om_table}).

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}
